Probabilistic Forecasting of Geomagnetic Indices using Gaussian Process Models

### 2017-08-07 revised-chapter.tex
Revisions [WIP]

\subsection{State of the art}

@@

\emph{Gaussian Processes} first appeared in machine learning research in \citet{Neal:1996:BLN:525544}, as the limiting case of Bayesian inference performed on neural networks with infinitely many neurons in the hidden layers. Although their inception in the machine learning community is recent, their origins can be traced back to the geo-statistics research community where they are known as \emph{Kriging} methods \citep{krige1951statistical}. In pure Mathematics, \emph{Gaussian Processes} have been studied extensively and their existence was first proven by Kolmogorov's extension theorem \citep{tao2011introduction}. The reader is referred to \citet{Rasmussen:2005:GPM:1162254} for an in depth treatment of Gaussian Processes in machine learning.

@@

For the success of a \emph{Gaussian Process} model an appropriate choice of kernel function is paramount. The symmetry and positive semi-definiteness of \emph{Gaussian Process} kernels implies that they represent inner-products between some basis function representation of the data. The interested reader is suggested to refer to \citet{Berlinet2004}, \citet{Scholkopf:2001:LKS:559923} and \citet{hofmann2008} for a thorough treatment of kernel functions and the rich theory behind them. Some common kernel functions used in machine learning are listed in Table \ref{table:kernel}.

@@

The quantities $l$ in the RBF, and $b$ and $d$ in the polynomial kernel are known as \emph{hyperparameters}. Hyperparameters give flexibility to a particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively. The process of assigning values to the \emph{hyperparameters} is crucial in the model building process and is known as \emph{model selection}.

@@

We selected OMNI data sections 00:00 January 3 2010 to 23:00 January 23 2010 and 20:00 August 5 2011 to 22:00 August 6 2011 for training the GP-AR and GP-ARX models. The first training data section consists of ambient fluctuations of $Dst$ while the second contains a geomagnetic storm. A set of 24 storm events listed in Table \ref{table:validationstorms} is kept aside as a validation data set for selecting model orders $p, p_v, p_b$ while a list of 63 storm events listed in table \ref{table:teststorms} is used for final model evaluation.

@@

Figures \ref{fig:CompareMae} and \ref{fig:CompareCC} show how the mean absolute error and coefficient of correlation as calculated on the validation set storm events of Table \ref{table:validationstorms}, vary with increasing model order for GP-AR and GP-ARX. The results are represented as box and whisker plots, in which a rectangle is drawn to represent the first and third quartiles, with a horizontal line inside to indicate the median value, outlying points are shown as dots while the whiskers indicate the smallest and largest non-outliers. In both cases, the predictive performance first improves and then stagnates or worsens with increasing model order.

@@

After choosing the best performing GP-AR and GP-ARX models, we calculate their performance on the test set of Table \ref{table:teststorms}. The results of these model evaluations are summarized in Table \ref{table:results}, the GP-AR and GP-ARX models improve upon the performance of the \emph{persistence model}.

>>>>

\subsection{Models \& Algorithms}

@@

\emph{Gaussian Processes} first appeared in machine learning research
in \citet{Neal:1996:BLN:525544}, as the limiting case of Bayesian
inference performed on neural networks with infinitely many neurons in
the hidden layers. Although their inception in the machine learning
community is recent, their origins can be traced back to the
geo-statistics research community where they are known as
\emph{Kriging} methods \citep{krige1951statistical}. In pure
Mathematics, \emph{Gaussian Processes} have been studied extensively
and their existence was first proven by Kolmogorov's extension theorem
\citep{tao2011introduction}. The reader is referred to
\citet{Rasmussen:2005:GPM:1162254} for an in depth treatment of
Gaussian Processes in machine learning. In space weather forecasting
literature \emph{Gaussian Process Regression} has been used for one
hour ahead prediction of the Dst index \citep{SWE:SWE20479}.

@@

For the success of a \emph{Gaussian Process} model an appropriate choice of kernel function is paramount. The symmetry and positive semi-definiteness of \emph{Gaussian Process} kernels implies that they represent inner-products between some basis function representation of the data. The interested reader is suggested to refer to \citet{Berlinet2004}, \citet{Scholkopf:2001:LKS:559923} and \citet{hofmann2008} for a thorough treatment of kernel functions and the rich theory behind them. Some common kernel functions used in machine learning are listed in table \ref{table:kernel}.

@@

The quantities $l$ in the \emph{Radial Basis Function kernel}, and $b$ and $d$ in the polynomial kernel are known as \emph{hyperparameters}. Hyperparameters give flexibility to a particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively. The process of assigning values to the \emph{hyperparameters} is crucial in the model building process and is known as \emph{model selection}.

@@

We selected OMNI data sections 00:00 January 3 2010 to 23:00 January 23 2010 and 20:00 August 5 2011 to 22:00 August 6 2011 for training the GP-AR and GP-ARX models. The first training data section consists of ambient fluctuations of $Dst$ while the second contains a geomagnetic storm. A set of 24 storm events listed in table \ref{table:validationstorms} is kept aside as a validation data set for selecting model orders $p, p_v, p_b$ while a list of 63 storm events listed in table \ref{table:teststorms} is used for final model evaluation.

@@

Figures \ref{fig:CompareMae} and \ref{fig:CompareCC} show how the mean absolute error and coefficient of correlation as calculated on the validation set storm events of table \ref{table:validationstorms}, vary with increasing model order for GP-AR and GP-ARX. The results are represented as box and whisker plots, in which a rectangle is drawn to represent the first and third quartiles, with a horizontal line inside to indicate the median value, outlying points are shown as dots while the whiskers indicate the smallest and largest non-outliers. In both cases, the predictive performance first improves and then stagnates or worsens with increasing model order.

@@

After choosing the best performing GP-AR and GP-ARX models, we calculate their performance on the test set of table \ref{table:teststorms}. The results of these model evaluations are summarized in table \ref{table:results}, the GP-AR and GP-ARX models improve upon the performance of the \emph{persistence model}.
